Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-12-11 inst/manuscript/manuscript.Rmd
typo

Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations. @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence. An observation is an outlier if it has an extreme response value relative to expectation. The response GLM residuals simply compare the observation to its fitted latent mean:
\begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align}
Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility. Deviance, Pearson, and standardized residuals are all different function of responses residuals that are appropriately scaled to behave more like response residuals from a standard linear model. For example, deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align}
where $\mathbf{d}$ is a vector of individual deviances. The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit. The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\text{y}_i = f^{-1}(\hat{\text{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized]. Deviance is often used as a fit statistic, whereby lower values of deviance imply a better model fit (given the observed data). Scaling the response residuals by the reciprocal square root of the variance function (for the appropriate response distribution) yields the Pearson residuals, and scaling the Pearson residuals by the reciprocal square root of one minus the leverage value (which we discuss next) yields the standardized residuals [@faraway2016extending].

>>>>

Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations. @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence. An observation is an outlier if it has an extreme response value relative to expectation. The response GLM residuals simply compare the observation to its fitted latent mean:
\begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align}
Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility. Deviance, Pearson, and standardized residuals are all different functions of response residuals that are appropriately scaled to behave more like response residuals from a standard linear model. For example, deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align}
where $\mathbf{d}$ is a vector of individual deviances. The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit. The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\text{y}_i = f^{-1}(\hat{\text{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized]. Deviance is often used as a fit statistic, whereby lower values of deviance imply a better model fit (given the observed data). Scaling the response residuals by the reciprocal square root of the variance function (for the appropriate response distribution) yields the Pearson residuals, and scaling the Pearson residuals by the reciprocal square root of one minus the leverage value (which we discuss next) yields the standardized residuals [@faraway2016extending].

### 2025-12-11 inst/manuscript/manuscript.tex
typo

Inspecting model diagnostics is an important step of the modeling
process that can yield valuable insights into model behavior and unusual
observations. \citet{montgomery2021introduction} contextualize three
components of unusual observations: outliers, leverage, and influence.
An observation is an outlier if it has an extreme response value
relative to expectation. The response GLM residuals simply compare the
observation to its fitted latent mean: \begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align} Because observations often have a unique support in a GLM
(e.g., only two possible response values for binary data) and the
variance of an observation generally depends on its mean, response
residuals lack some utility. Deviance, Pearson, and standardized
residuals are all different function of responses residuals that are
appropriately scaled to behave more like response residuals from a
standard linear model. For example, deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
The sum of the squared deviance residuals equals the sum of the elements
of \(\mathbf{d}\), known as the deviance of the model fit. The deviance
of the model fit quantifies twice the difference in log likelihoods
between the a saturated model that fits every observation perfectly
(i.e., \(\text{y}_i = f^{-1}(\hat{\text{w}}_i)\) for all \(i\)) and the
fitted model \citep{myers2012generalized}. Deviance is often used as a
fit statistic, whereby lower values of deviance imply a better model fit
(given the observed data). Scaling the response residuals by the
reciprocal square root of the variance function (for the appropriate
response distribution) yields the Pearson residuals, and scaling the
Pearson residuals by the reciprocal square root of one minus the
leverage value (which we discuss next) yields the standardized residuals
\citep{faraway2016extending}.

>>>>

Inspecting model diagnostics is an important step of the modeling
process that can yield valuable insights into model behavior and unusual
observations. \citet{montgomery2021introduction} contextualize three
components of unusual observations: outliers, leverage, and influence.
An observation is an outlier if it has an extreme response value
relative to expectation. The response GLM residuals simply compare the
observation to its fitted latent mean: \begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align} Because observations often have a unique support in a GLM
(e.g., only two possible response values for binary data) and the
variance of an observation generally depends on its mean, response
residuals lack some utility. Deviance, Pearson, and standardized
residuals are all different functions of response residuals that are
appropriately scaled to behave more like response residuals from a
standard linear model. For example, deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
The sum of the squared deviance residuals equals the sum of the elements
of \(\mathbf{d}\), known as the deviance of the model fit. The deviance
of the model fit quantifies twice the difference in log likelihoods
between the a saturated model that fits every observation perfectly
(i.e., \(\text{y}_i = f^{-1}(\hat{\text{w}}_i)\) for all \(i\)) and the
fitted model \citep{myers2012generalized}. Deviance is often used as a
fit statistic, whereby lower values of deviance imply a better model fit
(given the observed data). Scaling the response residuals by the
reciprocal square root of the variance function (for the appropriate
response distribution) yields the Pearson residuals, and scaling the
Pearson residuals by the reciprocal square root of one minus the
leverage value (which we discuss next) yields the standardized residuals
\citep{faraway2016extending}.
